---
author: "Gabriele Tomè, Toma Tebaldi"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmdformats::readthedown:
    css: custom_theme_report.css
    df_print: paged
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
params:
  doc_title: "ProTN report"
  description: NA
  readPD_files: NA
  readMQ_files: NA
  db_execution: NA
  file_input: NA
  batch_corr_exe: NA
  prot_boxplot: NA
  fc_thr: "0.75"
  pval_fdr: NA
  pval_thr: "0.05"
  pval_fdr_enrich: NA
  pval_enrich_thr: "0.05"
  overlap_size_enrich_thr: 5
  enrich_filter_term: NA
  enrich_filter_DBs: NA
  taxonomy: NA
  score_thr: NA
  dirOutput: NA
  filt_absent_value: "0"
---


```{r setup, message=F, warning=F, echo=F,results='hide'}
library(proTN)
library(stringi)
library(stringr)

doc_title <- params$doc_title
description <- params$description
readPD_files <- params$readPD_files
readMQ_files <- params$readMQ_files
db_execution <- params$db_execution
file_input <- params$file_input
batch_corr_exe <- params$batch_corr_exe
list_proteins <- stri_split(stri_replace_all(regex = " ",
                                             replacement = "",
                                             str = params$prot_boxplot), regex=",")

fc_thr <- as.double(params$fc_thr)
pval_fdr <- params$pval_fdr
pval_thr <- as.double(params$pval_thr)
pval_fdr_enrich <- params$pval_fdr_enrich
pval_enrich_thr <- as.double(params$pval_enrich_thr)
overlap_size_enrich_thr <- as.double(params$overlap_size_enrich_thr)
enrich_filter_term <- unlist(stri_split(stri_replace_all(regex = "\"|;|.",
                                                         replacement = "",
                                                         str = params$enrich_filter_term), regex=","))

enrich_filter_DBs <- params$enrich_filter_DBs
taxonomy <- params$taxonomy
score_thr <- params$score_thr
dirOutput <- params$dirOutput
filt_absent_value <- params$filt_absent_value

### Knit Tab for figure ----

in_tabs <- function(l, labels = names(l), level, knit = TRUE, close_tabset = FALSE) {                       
  if(is.null(labels)) {                               
    stop("labels are NULL, it is required not to be so that the tabs have proper names")             
  }                       
  names(l) <- labels      
  rmd_code <- lapply(seq_along(l), FUN = function(i) obj_to_rmd(l[[i]], name = names(l)[i], level = level + 1L))                     
  if(isTRUE(getOption("knitr.in.progress"))) {                 
    res <- knitr::knit(text = unlist(rmd_code), quiet = TRUE)  
    cat(res)                                                   
  } else {                
    if(!knit) {             
      cat(unlist(rmd_code))         
    } else {           
      return(l)       
    }                   
  }                      
  if(close_tabset) {                            
    cat(paste(get_section(level), "{.unlisted .unnumbered .toc-ignore .tabset}", "\n"))                    
  }                     
}                                                                                                                                      
get_section <- function(level) {  
  paste(rep("#", times = level), collapse = "")      
}                 

get_tabset <- function(obj) { 
  ifelse(inherits(obj, "list"), "{.tabset}", "")
}                                                                                                              

obj_to_rmd <- function(obj, parent_name = "l", name, level) {                          
  section_code <- sprintf("%s %s %s\n", get_section(level), name, get_tabset(obj))              
  if(!inherits(obj, "list")) {             
    rmd_code <- c("```{r, echo = FALSE}\n",  
                  sprintf("%s$`%s`\n", parent_name, name),
                  "```\n",                        
                  "\n")                    
  } else {               
    rmd_code <- c("\n",            
                  lapply(X = seq_along(obj),                         
                         FUN = function(i) obj_to_rmd(obj[[i]], sprintf("%s$`%s`", parent_name, name), names(obj)[i], level + 1L)))
  }                           
  return(c(section_code, rmd_code))    
} 

### Function to plot Enrichment category ----

resize_plot <- function(resizePlot, resizeHeight) {
  resizePlot <- resizePlot
  resizeHeight <- resizeHeight
  res <- rmarkdown::render("enrich_plot.Rmd", quiet = T)
  htmltools::includeHTML(res)
}

save(list=ls(), file = "tmp.RData")
```


---
title: `r if(doc_title==""){"ProTN report"}else{doc_title}`
---

`r description`

The following report summarizes the results obtained by the proTN integrative pipeline for proteomics analysis. Full results can be found in the output folder.

```{r print_readme, warning=FALSE, echo=F, message=FALSE}

text_readme<- c("Description of the output files")

```

## Summary of input parameters and thresholds

```{r print_settings, warning=FALSE, echo=F, message=FALSE, results='asis'}
cat(" \n \n")
#Print summary of filter and contrast design
if(readPD_files){
  cat("**Source files:** Proteome Discoverer")
  cat(" \n")
} else if(readMQ_files){
  cat("**Source files:** MaxQuant")
  cat(" \n")
} else if(readTMT_files){
  cat("**Source files:** TMT with Protein Discoverer")
  cat(" \n")
}
cat(" \n \n")
if(!is.null(batch_corr_exe)){
  cat("**Pipeline** \n ")
  cat(" \n")
  cat("- Execution of batch effect correction.")
}
cat(" \n \n ")
cat("**Differential expression** \n ")
cat(" \n")
cat(paste0("- Fold Change threshold: log2_FC > ",fc_thr," (+), log2_FC < -",fc_thr," (-) "))
cat(" \n")
cat(paste0("- Statistical significance threshold: p.val < ",pval_thr," "))
cat(" \n")
if(filt_absent_value != 0){cat(paste0("- Threshold of acceptable missing values for condition = ",filt_absent_value," "))}
cat(" \n \n ")
if(length(db_execution$enrichmnent_results)>0){
  cat("**Enrichment analysis** \n ")
  cat(" \n")
  cat(paste0("- Statistical significance threshold: p.val < ",pval_enrich_thr," "))
  cat(" \n")
  cat(paste0("- Overlap size threshold for enrichment analysis: OverlapSize > ",overlap_size_enrich_thr," "))
  cat(" \n \n ")
}
cat(" \n \n ")
if(length(db_execution$stringdb_res)>0){
  cat("**Network analysis** \n ")
  cat(" \n")
  cat("- Execution of PPI network analysis with STRINGdb.")
  cat(paste0("- Score threshold: ", score_thr,"\n"))
}
```

## Analysis

### Sample annotation
The sample table below annotates the samples of the study as in the **Sample_Annotation** file. Important columns for the analysis:

- **Sample**: sample name used in the results,
- **Condition**: condition of the sample used in study design and differential analyses

```{r print_sample, warning=FALSE, echo=F, message=FALSE, results='asis'}
cat(" \n")
kable(db_execution$proteome_data$c_anno,
      row.names = F,
      col.names = str_to_title(colnames(db_execution$proteome_data$c_anno)))
cat(" \n")
```


Raw abundances are determined from the area under the peak (m/z, time, number of events) for each PSM (Peptide-to-Spectrum Match).

Initial filtering removes the following peptides:

-   Low/missing abundance: PSMs not consistently present in all replicates of at least one condition are removed.
-   Contaminants.

After filtering, raw abundances are log2 transformed

```{r read_proteomics_MQ, warning=F, echo=F, eval=readMQ_files, results='asis'}
proteome_data <- read_proteomics(software = "MQ",
                                 folder = file_input,
                                 peptide_filename = "evidence")
```

```{r read_proteomics_PD, warning=F, echo=F, eval=readPD_files, results='asis'}
proteome_data <- read_proteomics(software = "PD",
                                 folder = file_input)
```

```{r preprocess_figure_1, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$generate_abundance)}
cat("#### Distributions of log2 raw abundances \n")
cat("Violin plot showing sample distributions of log2 raw abundances of peptides. \n ")
# Avalaible abundance. The red percentage will be imputed.
db_execution$generate_abundance
```

```{r preprocess_figure_2, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$generate_peptide_distribution)}
cat("#### Percentage of quantified peptides per sample \n")
cat("Barplot showing the percentage of peptides with available abundances for each sample. \n ")
# Number peptide per each protein
db_execution$generate_peptide_distribution
```

#### Imputation, normalization, peptide-to-protein summarization 

**Imputation** is performed on peptide and protein abundances with the Bioconductor package [PhosR](https://www.sciencedirect.com/science/article/pii/S221112472100084X). Round imputation is performed in absence of replicates. ProTN uses two function of PhosR for the imputation, the first impute the missing values for a peptide across replicates within a single condition, the second is a tail-based imputation approach as implemented in Perseus.

**Normalization:**

**Peptides** abundances are normalized with the equalMedianNormalization function from the Bioconductor package [DEqMS](<https://www.bioconductor.org/packages/release/bioc/html/DEqMS.html>) (equal medians in all samples). 

**Protein** normalized abundances were calculated with the medianSweeping function from the Bioconductor package [DEqMS](<https://www.bioconductor.org/packages/release/bioc/html/DEqMS.html>).



```{r normalized_plot_prot, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$generate_peptide_distribution)}
cat("# Distribution of protein normalized abundance \n")
cat("Violin plot showing sample distributions of log2 normalized and imputed abundances. \n ")
#Abundance per Protein
db_execution$protein_abundance_distribution
```

```{r normalized_plot_pep, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$generate_peptide_distribution)}
cat("# Distribution of peptide normalized abundance \n")
cat("Violin plot showing sample distributions of log2 normalized and imputed abundances. \n ")
#Abundance per Peptide
db_execution$peptide_abundance_distirbution
```

```{r mds_normalized_prot, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$protein_MDS)}
cat("# MDS based on protein normalized data \n ")
cat("Multidimensional scaling (MDS) takes a set of dissimilarities and returns a set of points such that the distances between the points are approximately equal to the dissimilarities. Classical multidimensional scaling is also known as principal coordinates analysis. \n \n ")
#Abundance per Protein
db_execution$protein_MDS
```

```{r mds_normalized_pep, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$peptide_MDS)}
cat("# MDS based on peptide normalized data \n")
cat("Multidimensional scaling (MDS) takes a set of dissimilarities and returns a set of points such that the distances between the points are approximately equal to the dissimilarities. Classical multidimensional scaling is also known as principal coordinates analysis. \n  \n ")
#Abundance per Peptide
db_execution$peptide_MDS
```

```{r pca_normalized_prot, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$protein_PCA)}
cat("# PCA based on protein normalized data \n")
cat("Principal component analysis (PCA) is performed on normalized Protein and Peptide abundances. \n \n ")
#Abundance per Protein
db_execution$protein_PCA
```

```{r pca_normalized_pep, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$peptide_PCA)}
cat("# PCA based on peptide normalized data \n")
cat("Principal component analysis (PCA) is performed on normalized Protein and Peptide abundances. \n \n ")
#Abundance per Peptide
db_execution$peptide_PCA
```


```{r boxplot_protein, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=(length(list_proteins[[1]])>0 & !is.null(db_execution$protein_boxplot))}
cat("#### Boxplot abundances of selected proteins \n")
cat("Normalized log2 abundances are displayed for proteins selected in the input. Abundances are grouped per condition. \n \n ")
cat("Selected proteins with available abundances: \n")
cat(paste(list_proteins[[1]], sep = ", "))
cat(" \n \n ")
db_execution$protein_boxplot
```

```{r heatmap_protein, fig.height=7, fig.width=7,results='asis', echo=FALSE, eval=(length(list_proteins[[1]])>0 & !is.null(db_execution$protein_heatmap))}
cat("#### Heatmap abundances of selected proteins \n")
cat("Normalized log2 abundances are displayed for proteins selected in the input. \n ")
cat("Selected proteins with available abundances: \n")
cat(paste(list_proteins[[1]], sep = ", "))
cat(" \n \n ")
db_execution$protein_heatmap
```



```{r differential_analysis, echo=FALSE, results='asis', eval=length(db_execution$differential_results)>0}
cat("## Differentially analysis \n")
# cat("Differentially expressed peptides are identified using the Limma Bioconductor package. \n ")
eval_single_Cond <- length(db_execution$differential_results)>0
```

```{r print_settings_diff_analysis, warning=FALSE, echo=F, message=FALSE, results='asis', eval=eval_single_Cond}
cat(" \n \n")
#Print summary of filter and contrast design
cat("**Thresholds used:** \n ")
cat(" \n")
cat(paste0("- Fold Change threshold: log2_FC > ",fc_thr," (+), log2_FC < -",fc_thr," (-) "))
cat(" \n")
cat(paste0("- Statistical significance threshold: p.val < ",pval_thr," "))
cat(" \n")

```

```{r print_settings_design, warning=FALSE, echo=F, message=FALSE, results='asis', eval=eval_single_Cond}
cat("### Differential expression contrasts \n \n")
cat("The following table contains the names and the formulas of the contrasts considered for differential expression analysis. \n \n")
cat("**Contrast design formula are:** ")
cat(" \n")
kable(data.frame("Name"=names(db_execution$formule_contrast), 
                 "Formula"=unlist(db_execution$formule_contrast)), 
      row.names = F)
cat(" \n")
```

```{r differentiated_protein, fig.height=3, fig.width=7, echo=FALSE, results='asis', eval=(eval_single_Cond & !is.null(db_execution$protein_differential_barplot))}
cat("#### Number of differentially expressed proteins for each comparison \n")
db_execution$protein_differential_barplot
```

```{r differentiated_peptide, fig.height=3, fig.width=7, echo=FALSE, results='asis', eval=(eval_single_Cond & !is.null(db_execution$peptide_differential_barplot))}
cat("#### Number of differentially expressed peptide for each comparison \n")
db_execution$peptide_differential_barplot
```

```{r volcano_protein, fig.height=6, fig.width=6, warning=FALSE, echo=FALSE, results='asis', eval=(eval_single_Cond & !is.null(db_execution$protein_vulcano))}
cat("#### Volcano plots of Proteins {.tabset} \n")
cat("Differentially expressed proteins are displayes according to the thresholds set in the input. \n ")
cat("\n - \"+\" up-regulated protein \n ")
cat("\n - \"-\" down-regulated protein \n ")
cat("\n - \"=\" invariant protein \n \n ")
cat("Top enriched proteins are labeled. \n")
cat(" \n \n")
in_tabs(db_execution$protein_vulcano, labels = names(db_execution$protein_vulcano), level = 4L)
cat(" \n \n")
```

```{r volcano_peptide, fig.height=6, fig.width=6, warning=FALSE, echo=FALSE, results='asis', eval=(eval_single_Cond & !is.null(db_execution$peptide_vulcano))}
cat("#### Volcano plots of Peptide {.tabset} \n")
cat("Differentially expressed peptide are displayes according to the thresholds set in the input. \n ")
cat("\n - \"+\" up-regulated peptide \n ")
cat("\n - \"-\" down-regulated peptide \n ")
cat("\n - \"=\" invariant peptide \n \n ")
cat("Top enriched peptide are labeled. \n")
cat(" \n \n")
in_tabs(db_execution$peptide_vulcano, labels = names(db_execution$peptide_vulcano), level = 4L)
cat(" \n \n")
```


```{r mds_differential_normalized_prot, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$protein_differential_MDS)}
cat("# MDS based on differentially expressed proteins \n")
cat("Multidimensional scaling (MDS) are performed on normalized abundances of differentially expressed Proteins. \n")
#Abundance per Protein
db_execution$protein_differential_MDS
```

```{r mds_differential_normalized_pep, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$peptide_differential_MDS)}
cat("# MDS based on differentially expressed peptides \n")
cat("Multidimensional scaling (MDS) are performed on normalized abundances of differentially expressed Peptides. \n")
#Abundance per Peptide
db_execution$peptide_differential_MDS
```

```{r pca_differential_normalized_prot, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$protein_differential_PCA)}
cat("# PCA based on differentially expressed proteins \n")
cat("Principal component analysis (PCA) are performed on normalized abundances of differentially expressed Proteins. \n")
#Abundance per Protein
db_execution$protein_differential_PCA
```

```{r pca_differential_normalized_pep, fig.height=5, fig.width=7,results='asis', echo=FALSE, eval=!is.null(db_execution$peptide_differential_PCA)}
cat("# PCA based on differentially expressed peptides \n")
cat("Principal component analysis (PCA) are performed on normalized abundances of differentially expressed  Peptides. \n")
#Abundance per Peptide
db_execution$peptide_differential_PCA
```

```{r enrichment_sms,message=F, warning=F,echo=F, eval=length(db_execution$enrichmnent_results)>0, results='asis'}
cat("## Functional enrichment analysis \n")
cat("Annotation enrichment analysis is performed using the web-based application [EnrichR](https://maayanlab.cloud/Enrichr/), against an updated collection of 98 databases covering Transcription, Pathways, Ontologies, Diseases, Drugs, Cell Types, Experimental Datasets. \n \n ")
cat("For each comparison enrichment is calculate over up-regulated proteins (UP), down-regulated proteins (DOWN), and their union (ALL). \n")
cat(" \n")
cat(paste0("- Statistical significance threshold: p.val < ",pval_enrich_thr," "))
cat(" \n")
cat(paste0("- Overlap size threshold for enrichment analysis: OverlapSize > ",overlap_size_enrich_thr," "))
cat(" \n")
def_val_height=  knitr::opts_chunk$get("fig.height")
run_enrich <- length(db_execution$enrichmnent_results)>0
```

```{r enrichment, echo=F, message=F, warning=F, eval = length(db_execution$enrichmnent_results)>0, results='asis'}

terms_enrich <- unlist(stri_split(stri_replace_all(regex = "\"|;|.",replacement = "",
                                                   str = enrich_filter_term), regex=","))
plots_down <- enrichment_figure(enr_df = db_execution$enrichmnent_results,
                                category = c("down","up"), 
                                enrich_filter_term = terms_enrich,
                                save=F)

#LOAD category EnrichR
dbs_default <- read_tsv("data/dbs_enrichR.txt", col_names = FALSE) %>% as.data.frame()
dbs_category <- dbs_default %>% split(f = as.factor(.$X2))
category_db <- lapply(dbs_category, function(x){filter(x, x[,1] %in% intersect(unique(db_execution$enrichmnent_results$anno_class), enrich_filter_DBs))})


cat("#### Top enrichments in selected annotation databases. {.tabset}  \n")
cat("Significant enrichments are displayed as filled dots, non-significant as empty dots. Point size is proportional to log2 Odds Ratio. \n ")
cat("The dot plots shows the top significant terms considering selected annotation databases. \n ")
cat("**Up-regulated and down-regulated proteins (UP,DOWN)** \n ")
cat(" \n \n")

for(db in names(plots_down)){
  cat(paste0("##### ",db," \n"))
  print(resize_plot(resizePlot = plots_down[[db]], resizeHeight =  max(min(20,uniqueN(db_execution$enrichmnent_results[anno_class %in% category_db[[db]][,1],
                                                                            "anno_name"])*0.3),3)))
  cat(" \n \n")
}
cat(" \n \n")
```

```{r string network, echo=F, message=F, warning=F, eval=length(db_execution$stringdb_res)>0, results='asis'}
cat("## Protein-protein interaction network analysis of differentially expressed proteins \n")
cat(paste0("Protein-protein interactions (PPIs) are crucial for understanding cell physiopathology. PPI analysis of differentially expressed proteins is based on the [STRING](https://string-db.org/) database.  Communities within the network are identified with the Clauset-Newman-Moore (CNM) algorithm. \n \n"))
# cat("For a better visualization of the network, only the top 500 differential proteins are considered for each contrast.")
db_execution$stringdb_res
```
